The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-05 paper.tex
lines => ln.

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, when \code{A} occurs, \code{B} is emitted (lines 
5--6) and both events become active, resulting in the invocation of \code{f()} 
and \code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (lines 3--6) that continuously 
awaits its identifying event (line 4), incrementing the value passed as 
reference (line 5).
A trail in parallel (lines 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (line 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (line 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (line 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test (line 11).

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (lines 3--6), in parallel 
with the caller (lines 8--11).
\label{lst.sub}
}
\end{figure}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (line 10) makes 
the periodic code to also execute immediately (line 9), and then in reactions 
to each \code{emit} request (line 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (line 2) and also to copy the condition test to the periodic 
code (line 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (lines 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (lines 4,7) and, according to the semantics 
of \CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (lines 5,8) occur in the same reaction to input 
\code{A} (lines 4,7).

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(line 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (line 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(line 15).
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

The example in the left of Figure~\ref{lst.impl.ast} has a corresponding $AST$, 
in the right of the figure, with the sets $I$ and $O$ for each node.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (lines 5,8 in the code) have an empty 
intersection of $I$ (lines 6,9 in the AST), hence, they do not conflict.
Note that although the accesses in lines 5,11 in the code (lines 6,11 in the 
AST) do have an intersection, they are not in parallel and are also safe.

@@

In the final generated code in \C, each trail label representing an entry point 
becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (lines 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(line 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in line 14).

@@

During the dispatch pass, if a trail executes and emits an internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to the current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
(lines 4--7) and proceeds to the code in \code{TRAIL-0} (lines 9--14), 
respecting the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
to await \code{EVT\_A} and also halts.

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19--23) and the radio transceiver 
(lines 25--36).

>>>>

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which serves the same purpose of \CEU's 
\code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions and programs cannot 
emit them, resulting in an intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, when \code{A} occurs, \code{B} is emitted (ln.  
5--6) and both events become active, resulting in the invocation of \code{f()} 
and \code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test (ln. 11).

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (ln. 2) and also to copy the condition test to the periodic code 
(ln. 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (ln. 4,7) and, according to the semantics of 
\CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 3);
the other periodically transmits the packet (ln. 5--18).
%
The periodic transmission is a loop that starts two other trails (ln. 6--17):
one handles the immediate retransmission request (ln. 7);
the other transmits the packet and waits for a confirmation (ln. 9--16).
%
%
The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (ln. 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(ln. 15).
%
At any time, the client may request a retransmission (ln. 7), which terminates 
the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
idle), and restarts the loop (ln. 5).
%
The client may also request to stop the whole protocol at any time (ln. 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

The example in the left of Figure~\ref{lst.impl.ast} has a corresponding $AST$, 
in the right of the figure, with the sets $I$ and $O$ for each node.
The event $.$ (dot) represents the ``boot'' reaction.
The assignments to \code{y} in parallel (ln. 5,8 in the code) have an empty 
intersection of $I$ (ln. 6,9 in the AST), hence, they do not conflict.
Note that although the accesses in ln. 5,11 in the code (ln. 6,11 in the AST) 
do have an intersection, they are not in parallel and are also safe.

@@

In the final generated code in \C, each trail label representing an entry point 
becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

During the dispatch pass, if a trail executes and emits an internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to the current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
(ln. 4--7) and proceeds to the code in \code{TRAIL-0} (ln. 9--14), respecting 
the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to 
await \code{EVT\_A} and also halts.

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (ln. 3) contains all definitions for the compiled \CEU 
program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (ln. 4) contains the main loop of \CEU pointing to the 
labels defined in the program.
The callback \code{Boot.booted} (ln. 6--11) is called by TinyOS on startup, so 
we initialize \CEU inside it (ln. 7).
If the \CEU program uses timers, we also start a periodic timer (ln. 8--10) 
that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (ln. 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (ln. 19--23) and the radio transceiver (ln.  
25--36).
